Vagueness \& Rationality

### 2018-09-01 springer/current/e1.tex
some consistency edits

Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Means for response times and proportion of borderline responses are given in Figure \ref{resultsC-exp-1}.

@@

\begin{description}
\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions actually led to significantly slower responses than crisp instructions, against Hypothesis 1: $\beta=0.058$, $se=0.013$, $t=4.55$, $p<0.001$. When the model was restricted to numeric-only instructions Vague instructions still led to significantly slower responses than crisp instructions $\beta=0.093$, $se=0.021$, $t=4.51$, $p<0.001$. When the model was restricted to verbal-only instructions Vague instructions tended to slow responses, but not significantly: $\beta=0.024$, $se=0.016$, $t=1.46$, $p=0.155$.
\item [Hypothesis 2 (Numeric/Verbal RT)] There was actually a significant difference between numeric and verbal instructions, with numeric instructions leading to longer responses than verbal instructions, against Hypothesis 2: $\beta0.265=$, $se=0.072$, $t=5.08$, $p<0.001$.
\item [Hypothesis 3 (Item RT)] Responses took longer as the levels of Item increased, supporting Hypothesis 3: $\beta=0.120$, $se=0.017$, $t=7.11$, $p<0.001$.
\end{description}

>>>>

30 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times and proportion of borderline responses are given in Figure \ref{resultsC-exp-1}.

@@

\begin{description}
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions actually led to significantly slower responses than crisp instructions, against Hypothesis 1: $\beta=0.058$, $se=0.013$, $t=4.55$, $p<0.001$. When the model was restricted to numeric-only instructions Vague instructions still led to significantly slower responses than crisp instructions $\beta=0.093$, $se=0.021$, $t=4.51$, $p<0.001$. When the model was restricted to verbal-only instructions Vague instructions tended to slow responses, but not significantly: $\beta=0.024$, $se=0.016$, $t=1.46$, $p=0.155$.
	\item [Test of Hypothesis 2 (Numeric/Verbal RT)] There was actually a significant difference between numeric and verbal instructions, with numeric instructions leading to longer responses than verbal instructions, against Hypothesis 2: $\beta0.265=$, $se=0.072$, $t=5.08$, $p<0.001$.
	\item [Test of Hypothesis 3 (Item RT)] Responses took longer as the levels of Item increased, supporting Hypothesis 3: $\beta=0.120$, $se=0.017$, $t=7.11$, $p<0.001$.
\end{description}

### 2018-09-01 springer/current/e2.tex
did a bit more on e1, still needs stats numbers imported from Rmds
some consistency edits

\marginnote{D-Exp-2}

The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., in whether it amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure \ref{Experiment1and2examplestimulus}). Table \ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

@@

\subsection{Hypotheses}

For Experiment 2, we hypothesised:

\begin{description}
	\item [Hypothesis 1] Vague instructions are easier for the reader than crisp ones (main effect of vagueness)
	\item [Hypothesis 2] Comparison is easier for the reader than matching (main effect of selection task)
	\item [Hypothesis 3] Effects of vagueness are different depending on whether the selection task is matching or comparison (interaction effect selection x vagueness).
\end{description}

\subsection{Experimental Method}

38 participants were recruited. We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table \ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded.

\subsection{Results}

We used the method proposed by \citet{Levy:MainEffectsInteractions} to test for the main effect of Vagueness in the presence of its higher-order interactions.

Response times were trimmed at 2.5 SD separately for each subject, leading to the loss of 204 trials (2.8\% of the trials). Condition means for the remaining (logged) RTs are plotted in Figure \ref{resultsD-exp-2}. A linear mixed model was constructed for the logged response times, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects.

@@

\begin{description}
	\item [Test of Hypothesis 1] Vague instructions were non-significantly easier for the reader than crisp instructions when vagueness was considered as a main effect ($\beta=-0.01$, $se=0.01$, $t=-0.4$, $p=0.7158$). 
	\item [Test of Hypothesis 2] There was a statistically significant main effect of selection task, with the comparison task speeding responses compared to the matching task ($\beta=0.16$, $se=0.03$, $t=6.2$, $p<0.0001$). 
	\item [Test of Hypothesis 3] Vagueness exerted effects in different directions for the comparison task and for the matching task: there was a significant interaction between vagueness and selection task ($\beta=0.13$, $se=0.03$, $t=4.2$, $p<0.0001$). 
Separate analyses were conducted testing for effects of vagueness at each level of the selection task.
Within the comparison task vagueness significantly speeded response times compared with crisp controls ($\beta=-0.07$, $se=0.02$, $t=-3.5$, $p<0.0012$). 
Within the matching task vagueness significantly \emph{slowed} response times compared with crisp controls ($\beta=0.06$, $se=0.02$, $t=2.9$, $p<0.0061$). 
\end{description}

In other words, the cost reduction account was wrong to predict significant main effect advantages for vagueness (although there was a non-significant trend in the \emph{direction} predicted by the cost reduction account), and wrong to predict that vagueness should be beneficial at each level of the selection task: however vagueness was significantly advantageous in the comparison task.

>>>>

% D_Exp_2

The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., whether the task amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure \ref{Experiment1and2examplestimulus}). We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table \ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded. Table \ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

@@

\subsection{Hypotheses (Experiment 2)}

For Experiment 2, we formulated the following hypotheses:

\begin{description}
	\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions should result in faster responses than crisp instructions.
	\item [Hypothesis 2 (Comparison/Matching RT)] Instructions that allow comparison should result in faster responses than instructions that necessitate matching.
	\item [Hypothesis 3 (Interaction)] The vagueness effect should differ according to whether the selection task is comparison or matching.
\end{description}

\subsection{Results (Experiment 2)}

38 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 204 trials (2.8\% of the trials). The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times are plotted in Figure \ref{resultsD-exp-2}.

@@

A linear mixed model was constructed for the logged response times, with sum-coded vagueness, selection task, and their interaction, and item as fixed effects, and per-participant intercepts and slopes for sum-coded vagueness, selection task, and their interaction, and item for random effects.

\begin{description} 
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.0057$, $se=0.0137$, $t=-0.42$, $p=0.678$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.676$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] comparison instructions resulted in faster responses than matching instructions, and the difference was significant ($\beta=0.1618$, $se=0.0255$, $t=6.34$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] Although Vagueness did not exert a significant main effect, Vagueness did exert effects in interactions with some other variables: the interaction between Vagueness and Selection task  was significant ($\beta=0.1306$, $se=0.0205$, $t=6.38$, $p<0.001$), suggesting that Vagueness speeded RTs in the comparison condition but slowed them down in the matching task. 
\end{description}

\subsection{Discussion (Experiment 2)}

The cost reduction account predicted that there should be a significant main effect of Vagueness such that responses would be faster for Vague intructions than for crisp instructions. We found that although there was a very small effect in that direction, the effect was not statistically significant. Models that differed only in the presence of Vagueness as a main effect were shown not to differ significantly in their explanatory value. However we did find that Vagueness exerted effects on other variables: Vagueness speeded RTs in the comparison task and slowed RTs in the matching task.

### 2018-09-01 springer/current/e3.tex
some consistency edits

\marginnote{E-Exp-3}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccll}
\hline\noalign{\smallskip}
Item & Target & Quantity & Selection & Crisp & Vague \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Small & Matching 	& the same number of dots as the target & NA \\ 
06:15:24 &  10 & Small & Matching 	& NA & about the same number of dots as the target \\ 
06:15:24 &  10 & Large & Comparison & more dots than the target & far more dots than the target \\ 
06:15:24 &  20 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
06:15:24 &  20 & Large & Matching 	& NA & about the same number of dots as the target \\ 
06:15:24 &  24 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 &  16 & Small & Matching 	& the same number of dots as the target & NA \\ 
16:25:34 &  20 & Small & Matching 	& NA & about the same number of dots as the target \\ 
16:25:34 &  20 & Large & Comparison & more dots than the target & far more dots than the target \\ 
16:25:34 &  30 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
16:25:34 &  30 & Large & Matching 	& NA & about the same number of dots as the target \\ 
16:25:34 &  34 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 &  26 & Small & Matching 	& the same number of dots as the target & NA \\ 
26:35:44 &  30 & Small & Matching 	& NA & about the same number of dots as the target \\ 
26:35:44 &  30 & Large & Comparison & more dots than the target & far more dots than the target \\ 
26:35:44 &  40 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
26:35:44 &  40 & Large & Matching 	& NA & about the same number of dots as the target \\ 
26:35:44 &  44 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 &  36 & Small & Matching 	& the same number of dots as the target & NA \\ 
36:45:54 &  40 & Small & Matching 	& NA & about the same number of dots as the target \\ 
36:45:54 &  40 & Large & Comparison & more dots than the target & far more dots than the target \\ 
36:45:54 &  50 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
36:45:54 &  50 & Large & Matching 	& NA & about the same number of dots as the target \\ 
36:45:54 &  54 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

\subsection{Hypotheses}

@@

\begin{description}
	\item [Hypothesis 1] Vague instructions are easier for the reader than crisp ones (main effect of vagueness)
	\item [Hypothesis 2] Comparison is easier for the reader than matching (main effect of selection)
	\item [Hypothesis 3] Effects of vagueness differ depending on whether selection is matching or comparison (interaction effect selection x vagueness, and focussed comparisons at each level of selection).
\end{description}

\subsection{Results}
40 volunteers participated. The results showed that vagueness was beneficial for comparison but detrimental for matching (the same as Experiment 2) even when no numbers were allowed in the instructions. Figure \ref{resultsE-exp-3} shows the means by condition. Our findings were as follows:

\begin{description}
	\item [Test of hypothesis 1] There was no significant main effect of vagueness ($\beta=-0.02$, $se=0.01$, $t=-1.5$, $p=0.1296$). 
	\item [Test of hypothesis 2] There was a main effect of selection, with comparison task instructions leading to faster responses than the matching task instructions ($\beta=0.18$, $se=0.02$, $t=10.4$, $p<0.0001$).  This effect was in the same direction as Experiment 3. 
	\item [Test of hypothesis 3] Vagueness did exert different effects depending on the selection task (main interaction effect of vagueness by selection $\beta=0.12$, $se=0.02$, $t=5.1$, $p<0.0001$). 
Separate analyses of the effect of vagueness were conducted for the comparison task and for the matching task using Bonferroni-adjusted significance thresholds. 
In the comparison task, vagueness resulted in faster response times ($\beta=-0.08$, $se=0.02$, $t=-4.3$, $p<0.0001$). 
In the matching task vagueness slowed response times ($\beta=0.05$, $se=0.01$, $t=3.7$, $p=0.0004$). 
These results are in the same direction as Experiment 3.
\end{description}

Once again, the cost reduction account was wrong to predict main effect advantages for vagueness, and wrong to predict that vagueness should be beneficial at each level of the selection task: however vagueness was advantageous in the comparison task.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ee3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 3}
\label{resultsE-exp-3}
\end{figure}

>>>>

%E_Exp_3

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\ 
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\ 
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

\subsection{Hypotheses (Experiment 3)}

@@

\begin{description}
	\item [Hypothesis 1 (Crisp/Vague RT)] Vague instructions are easier for the reader than crisp ones.
	\item [Hypothesis 2 (Comparison/Matching RT)] Comparison is easier for the reader than matching.
	\item [Hypothesis 3 (Interaction)] The effect of vagueness differs depending on whether the selection task mandated by the instructions is matching or comparison.
\end{description}

\subsection{Results (Experiment 3)}

40 volunteers participated. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 211 trials (2.8\% of the trials). The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times are plotted in Figure \ref{resultsE-exp-3}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ee3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 3 where all instructions were verbal}
\label{resultsE-exp-3}
\end{figure}

A linear mixed model was constructed for the logged response times, with sum-coded vagueness, selection task, and their interaction, and item as fixed effects, and per-participant intercepts and slopes for sum-coded vagueness, selection task, and their interaction, and item for random effects.

\begin{description}
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.015$, $se=0.0097$, $t=-1.58$, $p=0.119$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.118$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] Comparison instructions resulted in significantly faster responses than matching instructions ($\beta=0.176$, $se=0.0168$, $t=10.51$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] Although Vagueness did not exert a significant main effect, Vagueness did exert effects in its interaction with Selection task ($\beta=0.123$, $se=0.0166$, $t=7.42$, $p<0.001$), suggesting that Vagueness speeded response times in the comparison condition but slowed response times in the matching task. 
\end{description}

\subsection{Discussion (Experiment 3)}

The cost reduction account predicted that there should be a significant main effect of Vagueness such that responses would be faster for Vague intructions than for crisp instructions. We found that although there was a very small effect in that direction, the effect was not statistically significant. Models that differed only in the presence of Vagueness as a main effect were shown not to differ significantly in their explanatory value.

However, the results also showed that vagueness did exert an influence on reaction times according to whether the task was comparison or matching: vagueness was beneficial for comparison and detrimental for matching (the same as Experiment 2) even when no numbers were allowed in the instructions.
